Submodular Optimization for Control of Prosumer Networks

### 2016-04-13 Article.tex
update

Several studies highlighted the need for increasing the storage capacity of the system. Some of them even consider electric vehicles as moving capacities that could be used for frequency regulation. In the case of fixed storage devices, the locations where they should be installed is an important question since it impacts the performances of the system. In a very static and top-down architecture, where loads and generators are fixed, this question is far from trivial. But in a smart grid scenario with bi-directional flows and agents that produce and consume depending on weather conditions, this becomes even more challenging. In this direction, \cite{Gkatzikis2015} studies the placement of storage equipments within microgrids in a collaborative scheme. The authors solve an interesting optimization problem mixing placement-dimensioning of storage with utilization. Moreover, \cite{Gkatzikis2015} uses a Nash bargaining framework to determine how prosumers should share the costs and benefits of the storages.

@@

In this paper, we explore the use of storage in a network of prosumers whose power outputs are susceptible to change due to various external causes. In such a scenario with a high penetration of renewables, perturbations in the power distributions of the prosumers are very likely. Re-balancing the production and the consumption is then necessary and requires that we have the capability to control the grid's dynamic. Control of smart grid systems has been recently studied in \cite{Farraj2015}, where the authors use linear-quadratic optimal control theory in order to compute DER outputs. Control inputs rely on information collected by sensors and phasor measurement units and a communication network is used for sharing the information. The performances are then studied under practical limitations such as latency, sampling rate, or signal-to-noise ratio.

The present work differs in that we consider the control inputs as the actions of the storages on the grid dynamics. We are given a network of prosumers, and we have to find both the number and locations of the storages that should be deployed. Once this choice is made, we have to stick with it afterwards whatever the perturbations that may occur. We are thus looking for the smallest set of storages, as well as their locations, that will require a priori small control energy. Since we do not have any prior on the future perturbations, we minimize the average control energy required to move the system around the state space.

@@

Optimal control theory can be used to compute the control inputs needed to restaure synchrony. However, there is no guarantee that these inputs will respect the physical constraints of the system. Electrical lines have indeed maximum capacities, and storages are also limited with bounded rates of charge/discharge which limit the possible inputs. We propose in this paper to incorporate these constraints in the model such that the control inputs remain realistic. Furthermore, we propose an optimization method that uses submodular set functions in order to find both the number of storages and their locations such that the average control energy required to maintain stability in this constrained system is minimized.

@@

This paper is organized as follows, section \ref{sec:control} presents recent advances about control in networks, section \ref{sec:The_Model} introduces the oscillator model designed for simplifying the grid's dynamic and derives the constraints on the control inputs. In section \ref{sec:control_and_submodularity} we propose an optimization method for finding the subset of driver nodes such that the average control energy is optimized. Finally, in section \ref{sec:Results}, we show some results and validation.

@@

In this section, we introduce some key notions related to the controllability of complex networks. We consider a graph $G=(V,E)$ of N nodes, where V is the vertex set and E is the edge set. The topology of this network can be encoded through the adjacency matrix M, which elements $m_{ij} \in \{0,1\}$ indicate whether i and j are connected. If the edges are weighted we can set $m_{ij} = w_{ij}$ where $w_{ij}$ is the weight of edge$(i,j)$, and zero if there is no edge. Graphs are often used to model complex interactions inside a population. Lets attribute to each node i the variable $ x_i $ that represents the state of node i. Depending on the model, this state can represent an opinion, a physical quantity, a probability and so on. If there is an edge between i and j, we might expect some interaction that would eventually change the values of $x_i$ and $x_j$. Such a dynamic is usually described with a system of differential equations. If we assume a linear first order dynamic, this can be written as $ \dot{X}(t) = A X(t) $, where $X(t)=\{x_0(t),...,x_{N-1}(t)\}$ is the vector of node states at time t, and matrix A encodes both the dynamics and the topology of the system (A can be, among other, the weighted adjacency matrix of the underlying graph).

@@

It was shown by Kalman in 1963 that a linear system $(A,B)$ is controllable if the controllability matrix $ C = [B,AB,A^2B,...,A^{N-1}B] $ has full rank, that is, $rank(C) = N$. Nevertheless, when studying the controllability of non trivial networks, using the Kalman's criterion is not a simple task since matrix C becomes huge. Furthermore, because $A^k\longrightarrow 0$ when $k\longrightarrow \infty $, numeric precision becomes quickly an issue when computing the rank of C. A different approach to study controllability in complex network was introduced later on by Lin \cite{1100557}. This relates the search of the minimum driver set to a maximum matching in a bipartite graph, which can be done with the Hopcroft-Karp algorithm in $ O(|E|\sqrt{|V|}) $ in the worst case. This requires that the system $(A,B)$ is a structured system, meaning that all elements in A and B are either fixed zeros or independant free parameters. In such a situation, structural control is extremely powerfull since it enables us to discuss a network's controllability relatively quickly even if we do not know the exact weights of the edges. Note however that the independance of the parameters is a key assumption. If A is the adjacency matrix of an undirected network for instance, independance is not verified (since A is symetric). In this case, one should use different tools to study controllability \cite{Yuan2014}.

@@

Until now, we stayed very general on the state variables and the dynamics of the system. In the following section, we use a coupled oscillators model as to encode the grid dynamics in some equation of the form $\dot{X}=AX$.

@@

In this section, we explain the method that we used in order to find the driver nodes for the grid's dynamic (eq. \ref{eq:final_equation}). We first need to introduce submodular set functions and explain how their maximization can be achieved in reasonable time. More information about submodularity can be found in \cite{Krause2014}.

@@

@@

Generally speaking, finding the set $ S_k $ of size k that maximizes a set function F is a difficult problem because the number of sets grows exponentially with the number of nodes. Therefore complete enumeration and evaluation is only a feasible solution on very small examples. Nevertheless, if the set function is submodular, a simple greedy heuristic returns a solution $ S_k^{\star} $ such that, in the worst case, $ \frac{F(S_k^{\star})}{F(S_k^{OPT})} \sim 63\%$ (where $ S_k^{OPT}$ is the optimum set of size k) \cite{Krause2014}. This heuristic starts with a set S (possibly empty) and iteratively adds the element i that exhibits the highest marginal gain : $ F(S \cup \{i\}) \geq F(S \cup \{j\})\ \forall j $.

@@

Since the control energy is related to $ W^{-1} $, systems with "large" W will tend to be controlled with low energy. Still, this notion of "large" W is not very accurate and we need to quantify it by taking some metric based on W. We already mentioned the rank of W as being the dimension of the controllable subspace, but it is also known that the trace of W and $W^{-1}$ give information about average controllability and average control energy \cite{Summers2014}. In such a situation, we can build a set function that, given a set of driver nodes S, returns the value of one of these metrics. Indeed, given A and S, we can obtain the system dynamics $(A,B_S)$ and compute the gramian $W_S$. In other word, such a function would quantify the ability of a given set of nodes to contol the system on average.

A key point, demonstrated in \cite{Summers2014}, is that the set function $ F:S \longrightarrow Tr[W_S] $ is modular and the two functions $F:S \longrightarrow Tr[W^{-1}_S] $ ans $ F:S \longrightarrow rank[W_S] $ are submodular. This nice result enables us to look for the driver node set that optimizes some notion of the average controllability using a simple greedy heuristic with a worst case guarantee.

@@

\section{Conclusion}
\label{sec:Conclusion}
In this paper, we considered the case of networks of prosumers which can behave as generators or loads depending on weather conditions. We modeled the grid as a coupled oscillators network and approximated the dynamics with a system of second order linear differencial equations. We seek to control this dynamic when small perturbations occur by placing storage devices at a subset of the nodes. We believe that the energy that should be injected or absorbed from the grid by these controllers should be minimized. Given the diversity of possible situations, we seek the subset of nodes that, on average, necessitates minimum control energy. Based on the gramian matrix of the system and submodular set functions, we show how to find driver sets that use low energy on average and respect the system constraints.

>>>>

Several studies highlighted the need for increasing the storage capacity of the system. Some of them even consider electric vehicles as moving capacities that could be used for frequency regulation. In the case of fixed storage devices, the locations where they should be installed is an important question since it impacts the performances of the system. In a very static and top-down architecture, where loads and generators are fixed, this question is far from trivial. But in a smart grid scenario with bi-directional flows and agents that produce and consume depending on weather conditions, this becomes even more challenging. In this direction, \cite{Gkatzikis2015} studies the placement of storage equipments within microgrids in a collaborative scheme. The authors solve an interesting optimization problem mixing placement-dimensioning of storage with utilization. Moreover, \cite{Gkatzikis2015} uses a Nash bargaining framework to determine how prosumers should share the costs and benefits of the storages. However, \cite{Gkatzikis2015} does not take into account the grid dynamic as well as its controllability. Control of smart grid systems has been recently studied in \cite{Farraj2015}, where the authors use linear-quadratic optimal control theory in order to compute DER outputs. Control inputs rely on information collected by sensors and phasor measurement units and a communication network is used for sharing the information. The performances are then studied under practical limitations such as latency, sampling rate, or signal-to-noise ratio. The objective of \cite{Farraj2015} is the control of the system when perturbations occur. The control signals are applied to the DER, meaning that \cite{Farraj2015} does not consider a system where storage would be present and used for control.

@@

In this paper, we explore the use of storage in a network of prosumers whose power outputs are susceptible to change due to various external causes. In such a scenario with a high penetration of renewables, perturbations in the power distributions of the prosumers are very likely. Re-balancing the production and the consumption is then necessary and requires that we have the capability to control the grid's dynamic. The present work differs from \cite{Gkatzikis2015} and \cite{Farraj2015} in that we study the placement of the storages from a controllability perspective in a stochastic context due to DER. Moreover, we aim at models and methods that could be scallable to some extent.

We are given a network of prosumers, and we have to find both the number and locations of the storages that should be deployed. Once this choice is made, we have to stick with it afterwards whatever the perturbations that may occur. We are looking for the smallest set of storages, as well as their locations, that will require small control energy. Since we do not have any prior on the future perturbations, we minimize the average control energy required to move the system around the state space.

@@

Optimal control theory can be used to compute the control inputs needed to restaure synchrony. However, there is no guarantee that these inputs will respect the physical constraints of the system. Electrical lines have indeed maximum capacities, and storages are also limited with bounded rates of charge/discharge which limit the possible inputs. We incorporate these constraints in the model such that the control inputs remain realistic. Furthermore, we propose an optimization method that uses submodular set functions in order to find both the number of storages and their locations such that the average control energy required to maintain stability in this constrained system is minimized.

@@

This paper is organized as follows, section \ref{sec:control} presents recent advances about control in networks, section \ref{sec:The_Model} introduces the oscillator model of the grid's dynamic and derives the constraints on the control inputs. In section \ref{sec:control_and_submodularity} we propose the optimization method for finding the subset of driver nodes. Finally, in section \ref{sec:Results}, we show performance results and validation.

@@

In this section, we introduce some key notions related to the controllability of complex networks. We consider a graph $G=(V,E)$ of N nodes, where V is the vertex set and E is the edge set. The topology of this network can be encoded through the adjacency matrix M, which elements $m_{ij} \in \{0,1\}$ indicate whether i and j are connected. If the edges are weighted we can set $m_{ij} = w_{ij}$ where $w_{ij}$ is the weight of edge$(i,j)$, and zero if there is no edge. Graphs are often used to model complex interactions inside a population. Let us attribute to each node i the variable $ x_i $ that represents the state of node i. Depending on the model, this state can represent an opinion, a physical quantity, a probability and so on. If there is an edge between i and j, we might expect some interaction that would eventually change the values of $x_i$ and $x_j$. Such a dynamic is usually described with a system of differential equations. If we assume a linear first order dynamic, this can be written as $ \dot{X}(t) = A X(t) $, where $X(t)=\{x_0(t),...,x_{N-1}(t)\}$ is the vector of node states at time t, and matrix A encodes both the dynamics and the topology of the system (A can be, among other, the weighted adjacency matrix of the underlying graph).

@@

It was shown by Kalman in 1963 that a linear system $(A,B)$ is controllable if the controllability matrix $ C = [B,AB,A^2B,...,A^{N-1}B] $ has full rank, that is, $rank(C) = N$. Nevertheless, when studying the controllability of non trivial networks, using the Kalman's criterion is not a simple task since matrix C becomes huge. Furthermore, because $A^k\longrightarrow 0$ when $k\longrightarrow \infty $, numeric precision becomes quickly an issue when computing the rank of C. A different approach to study controllability in complex network was introduced later on by Lin \cite{1100557}. This relates the search of the minimum driver set to a maximum matching in a bipartite graph, which can be done with the Hopcroft-Karp algorithm in $ O(|E|\sqrt{|V|}) $ in the worst case. This requires that the system $(A,B)$ is a structured system, meaning that all elements in A and B are either fixed zeros or independant free parameters. In such a situation, structural control is extremely powerfull since it allows to discuss network control easily even if the exact weights of the edges are unknown. Note however that the independance of the parameters is a key assumption. If A is the adjacency matrix of an undirected network for instance, independance is not verified (since A is symetric). In this case, one should use different tools to study controllability \cite{Yuan2014}.

@@

@@

In this section, we explain the method that we used in order to find the driver nodes for the grid's dynamic (eq. \ref{eq:final_equation}).

Since the control energy is related to $ W^{-1} $, systems with "large" W will tend to be controlled with low energy. Still, this notion of "large" W is not very accurate and we need to quantify it by taking some metric based on W. We already mentioned the rank of W as being the dimension of the controllable subspace, but it is also known that the trace of W and $W^{-1}$ give information about average controllability and average control energy \cite{Summers2014}. In such a situation, we can build a set function that, given a set of driver nodes S, returns the value of one of these metrics. Indeed, given A and S, we can obtain the system dynamics $(A,B_S)$ and compute the gramian $W_S$. In other word, such a function would quantify the ability of a given set of nodes to contol the system on average. The objective would then be to find the set $S_k^{\star}$ of size k that maximizes this function.

A key point, demonstrated in \cite{Summers2014}, is that the set function $ F:S \longrightarrow Tr[W_S] $ is modular and the two functions $F:S \longrightarrow Tr[W^{-1}_S] $ ans $ F:S \longrightarrow rank[W_S] $ are submodular. As we will see in the next section, this nice result enables us to look for the driver node set that optimizes the average control energy with a simple greedy heuristic that also provides a worst case guarantee.

@@

We introduce here submodular set functions and explain how their maximization can be achieved in reasonable time. More information about submodularity can be found in \cite{Krause2014}.

@@

Generally speaking, finding the set $ S_k^{\star} $ of size k that maximizes a set function F is a difficult problem because the number of sets grows exponentially with the number of nodes. Therefore complete enumeration and evaluation is only a feasible solution on very small examples. Nevertheless, if the set function is submodular, a simple greedy heuristic returns a solution $ S_k^{\star} $ such that, in the worst case, $ \frac{F(S_k^{\star})}{F(S_k^{OPT})} \sim 63\%$ (where $ S_k^{OPT}$ is the optimum set of size k) \cite{Krause2014}. This heuristic starts with a set S (possibly empty) and iteratively adds the element i that exhibits the highest marginal gain : $ F(S \cup \{i\}) \geq F(S \cup \{j\})\ \forall j $.

@@

@@

\section{Conclusion}
\label{sec:Conclusion}
In this paper, we considered the case of networks of prosumers which can behave as generators or loads depending on weather conditions. We believe that the originality of this work was to study the storage placement in such a stochastic context through the optimal control theory. This approach enabled us to take the grid's dynamic and its physical constraints into account. Because perturbations could be of any kind, and because we only place the storages once, we proposed an optimization method, that rely on submodular set functions and the gramian matrix, for minimizing the average energy that should be injected or absorbed by the storage equipments.
